On The Subspace Learning for Network Attack Detection

### 2019-06-15 docs/thesis/chapters/4_m_rpca/m_rpca.tex
fixing

Some widely adopted algorithms for anomaly detection assume a Gaussian distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network, leon2017probability}.

Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. Leon-Garcia \cite{leon2017probability} also argues that Pareto distribution has been found to capture the behavior of many quantities of interest in the study of Internet behavior. Moreover, Benson \emph{et al.}  \cite{benson2010network} observes that the lognormal distribution is the best fit to model arrival processes in a data center.

These findings show that the skeweness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data. These findings can also highlight opportunities to  exploit the skeweness and heavy-tailed distributions to obtain accurate classifiers for network anomaly detection.

@@

Therefore, we propose the m-RPCA, which is an approach based on distances of moments computed from robust subspace learned by Robust Principal Component Analysis (RPCA), in order to detect anomalies from skewed data and network traffic. The proposed approach relies on a robust subspace computed from supposed normal observations, for estimating the moments to be used for distance analysis. The anomaly detection from new observations evaluate the Mahalanobis distance between the robust moments and the moments of new observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

@@

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Considering the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\bar{\pmb{x}}$ and covariance $\bar{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and new observations can reveal anomalies.

We propose an approach based on the distances of moments computed from learned robust subspace and new observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed normal traffic for estimating the moments (mean, skewness and kurtosis). Thus, the anomaly detection for new observations should evaluate the Mahalanobis Distance between the robust moments and the moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated moments.

@@

The CTU-13 data set originally contains the following features for each flow:

\begin{itemize}
	\item Start Time;
	\item End Time;
	\item Duration;
	\item Source IP Address;
	\item Source Port;
	\item Direction;
	\item Destination IP Address;
	\item Destination Port;
	\item State of TCP flags;
	\item Destination Type of Service;
	\item Source Type of Service;
	\item Total number of Packets;
	\item Total number of Bytes;
\end{itemize}

@@

This section describes the proposed approach for network attack detection by means of a distance analysis between moments computed from a robust subspace and moments of new observations of network traffic.

Robust subspace learning can be defined as the decomposition of a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, with outliers or noise. We propose to learn a robust subspace $\pmb{L}$ from the normal traffic $\pmb{X}^s$ for computing $\pmb{L}^s$, $\pmb{S}^s$ and the robust moments, i.e. the mean $\bar{\pmb{x}}$, skewness $\bar{\pmb{s}}$ and kurtosis $\bar{\pmb{k}}$, in order to evaluate the distance $\pmb{d}$ between moments of new observations $\pmb{X}^c$ and moments computed from the learned robust subspace, in order to classify as anomalous the observations with the largest distances.

@@

We propose to compute the mean $\bar{\pmb{x}}$, the skewness $\bar{\pmb{s}}$, the kurtosis $\bar{\pmb{k}}$ and the covariance matrix $\bar{\pmb{\Sigma}}$ from the robust subspace $\pmb{L}^s$, after to minimize the Equation \ref{eq:4.06} for $\pmb{X}^s$. We also propose to compute the Mahalanobis Distance (MD) for detecting anomalies between new observations and the moments calculated from a robust subspace computed by RPCA.

@@

The m-RPCA can also be adopted as unsupervised algorithm if the lines 1 and 2 of the Algorithm \ref{alg:4.01} are changed to substitute $\pmb{X}^s$ by $\pmb{X}^c$. Hence, the robust subspace is learned from contaminated data and used for comparing the distance between moments from robust estimate and moments of contaminated observations. It is important to note that this unsupervised approach requires the computational cost of computing new subspace learning for any new set of observations.

@@

The availability of labeled data is a challenging concern in real world problems of anomalous detection, where anomalies are rare or even unknown events. Considering that RPCA has already been adopted to isolate outliers from training data \cite{zhou2017anomaly}, we also propose to evaluate the m-RPCA for a semi-supervised approach based on training from contaminated data, in order to evaluate the impact that contaminated data for robust subspace learning can cause in the anomaly detection results.

Therefore, we propose to evaluate the following approaches for m-RPCA: Semi-supervised; Contaminated Semi-supervised; and Unsupervised.

@@

We finally evaluate the unsupervised approach, that relies on the test data sets $\pmb{Y}_g^c$, $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$ for robust subspace learning, and then classify the results according to the distance between moments of the contaminated data $\pmb{Y}_g^c$, $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$, and the moments computed from the learned robust subspace.

@@

This experiment consider only the semi-supervised approach due to results of simulated experiments on simulated data set to be present in Subsection \ref{sec:4_simulated_result}, that shows better results for the semi-supervised approach and highlight that the this approach can obtain good results even when trained with contaminated data set.

>>>>

Some widely adopted algorithms for anomaly detection assume a Gaussian or symmetric distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network, leon2017probability}.

Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. Leon-Garcia \cite{leon2017probability} also argues that Pareto distribution has been found to capture the behavior of many quantities of interest in the study of Internet behavior. Moreover, Benson \emph{et al.}  \cite{benson2010network} observes that the Lognormal distribution is the best fit to model arrival processes in a data center.

These findings show that the skewness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data. These findings can also highlight opportunities to  exploit the skewness and heavy-tailed distributions to obtain accurate classifiers for network anomaly detection.

An outlyingness-approach based on a robust estimator of skewness, combined with robust estimators of location and scale, can be able to flag the outlying measurements. When the same methodology would be used with non-robust estimators of location, scale and skewness, the outlyingness-values would be affected by the outliers such that the outlying group could be masked \cite{hubert2009robustskewed}.

@@

Therefore, we propose the m-RPCA, which is an approach based on distances between moments computed from a robust subspace learned by Robust Principal Component Analysis (RPCA) and contaminated observations, in order to detect anomalies from skewed data and network traffic. The proposed approach relies on a robust subspace computed from supposed normal observations, for estimating the moments to be used for distance analysis. The anomaly detection from contaminated observations evaluate the Mahalanobis distance between the robust moments and new contaminated observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

@@

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Taking into account the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\bar{\pmb{x}}$ and covariance $\bar{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and contaminated observations can reveal anomalies.

We propose an approach based on the distances between moments computed from learned robust subspace and contaminated observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed normal traffic for estimating the moments (mean, skewness and kurtosis). Thus, the anomaly detection for contaminated observations should evaluate the Mahalanobis Distance between the robust moments and contaminated observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated observations.

@@

The CTU-13 data set originally contains the following features for each flow: Start Time, End Time, Duration, Source IP Address, Source Port. Direction, Destination IP Address, Destination Port, State of TCP flags, Destination Type of Service, Source Type of Service, Total number of Packets, Total number of Bytes.

@@

This section describes the proposed approach for network attack detection by means of a distance analysis between moments computed from a robust subspace and contaminated observations of network traffic.

Robust subspace learning can be defined as the decomposition of a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, with outliers or noise. We propose to learn a robust subspace $\pmb{L}$ from the normal traffic $\pmb{X}^s$ for computing $\pmb{L}^s$, $\pmb{S}^s$ and the robust moments, i.e. the mean $\bar{\pmb{x}}$, skewness $\bar{\pmb{s}}$ and kurtosis $\bar{\pmb{k}}$, in order to evaluate the distance $\pmb{d}$ between contaminated observations $\pmb{X}^c$ and moments computed from the learned robust subspace, in order to classify as anomalous the observations with the largest distances.

@@

We propose to compute the mean $\bar{\pmb{x}}$, the skewness $\bar{\pmb{s}}$, the kurtosis $\bar{\pmb{k}}$ and the covariance matrix $\bar{\pmb{\Sigma}}$ from the robust subspace $\pmb{L}^s$, after to minimize the Equation \ref{eq:4.06} for $\pmb{X}^s$. We also propose to compute the Mahalanobis Distance (MD) for detecting anomalies between contaminated observations and the moments calculated from a robust subspace computed by RPCA.

@@

The m-RPCA can also be adopted as unsupervised algorithm if the lines 1 and 2 of the Algorithm \ref{alg:4.01} are changed to substitute $\pmb{X}^s$ by $\pmb{X}^c$. Hence, the robust subspace is learned from contaminated data and used for comparing the distance between moments from robust estimate and contaminated observations. It is important to note that this unsupervised approach requires the computational cost of computing new subspace learning for any new set of observations.

@@

The availability of labeled data is a challenging concern in real world problems of anomalous detection, where anomalies are rare or even unknown events. Considering that RPCA has already been adopted to isolate outliers from training data \cite{zhou2017anomaly}, we also propose to evaluate the m-RPCA for a contaminated semi-supervised approach based on training from contaminated data, in order to evaluate the impact that contaminated data for robust subspace learning can cause in the anomaly detection results.

Therefore, we propose to evaluate the following approaches for m-RPCA: semi-supervised; contaminated semi-supervised; and unsupervised.

@@

We finally evaluate the unsupervised approach, that relies on the test data sets $\pmb{Y}_g^c$, $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$ for robust subspace learning, and then classify the results according to the distance between the contaminated data $\pmb{Y}_g^c$, $\pmb{Y}_p^c$ and $\pmb{Y}_l^c$, and the moments computed from the learned robust subspace.

@@

This experiment consider only the semi-supervised approach due to results of simulated experiments on simulated data set presented in Subsection \ref{sec:4_simulated_result}, that shows better results for the semi-supervised approach and highlight that the this approach can obtain good results even when trained with contaminated data set.
